Controlled Natural Languages and Robotics

### 2018-07-07 paper/paper.tex
paper progress

@@

\section{Specification}
To summarize: at hand we have a parameterized system with $n$ process and $f$ number of faults. We want to show \textit{Floodset} is correct regardless of the parameters. It is trivial to verify for a small number of $n,\ f$, but the question is ``does \textit{Floodset} work for any assignment of $n,\ f, n \lt f$''.
To answer this question we are using the theorem proofer Ivy; first encoding the algorithm in the Ivy language and then constructing inductive invariants to proof the correctness.

@@

>>>>

\lstset{basicstyle=\small}

@@

\section{Specification}
To summarize: at hand we have a parameterized system with $n$ process and $f$ number of faults. We want to show \textit{Floodset} is correct regardless of the parameters. It is trivial to verify for a small number of $n,\ f$, but the question is ``does \textit{Floodset} work for any assignment of $n,\ f,\ n < f$''.
To answer this question we are using the theorem proofer Ivy; first encoding the algorithm in the Ivy language and then constructing inductive invariants to proof the correctness.

The goal of this section is to walk through the specification step by step, to give the reader an understanding of the paradigms relevant to automatic verification and Ivy in particular. The complete code can be found here \cite{github}.

\subsection{Modeling strategy}

\subsubsection{Modularity}
When verifying an algorithm like \textit{Floodset} in Ivy, an algorithm designer would most likely prefer to be able to translate their original algorithm to the Ivy language in as close to a one-to-one fashion as possible. However, this is not quite convenient for \textit{Floodset} and the framework it operates in.
To separate system and nodes one would have to rely heavily on iterations. E.g. for \textit{Step 1} the system would iterate over all nodes and let each node take care of sending it's own message.

\begin{mdframed}[backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=15, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
action round = {
	var iter := node.t.iter.create(0);
  while ~iter.is_end
	{
 	 call send_messages(iter.val)
	 iter := iter.next;
	}
}
action send_messages(n:node.t) = {
  node.message(n, V) := node.w(n, V);
  node.incomming_message(O, n) := true;
}

export round
\end{lstlisting}
\end{mdframed}

However, the code above does not actually execute an iteration over all nodes as one would expect. Looking at an Ivy trace we see what really happens:

\begin{mdframed}[backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=15, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
message := *;
incomming_message := *;
\end{lstlisting}
\end{mdframed}

Ivy non-deterministically assigns to all relations involved in the loop, instead of actually executing the loop.  This is because proofing correctness of loops can only be done through inductive loop invariants specified by a human. Ivy therefore allows specifying invariants together with a loop.
\begin{mdframed}[backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=15, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
  while ~iter.is_end
  invariant node.message(N,V) = node.w(N,V)
  invariant N<=iter.val -> incomming_message(O, N)
	{
    . . .
	}
\end{lstlisting}
\end{mdframed}

These invariants must hold before the loop and after every iteration. The following is a trace of the execution of the loop with the specified invariants. First they are asserted, then the relations are set to non-deterministic values and finally an assume is used to restrict control to values for the relation that fulfil the invariants.

\begin{mdframed}[backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=15, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
assert message(N,V) = w(N,V)
assert N:t <= t.iter.val(loc:iter) -> incomming_message(O,N)
incomming_message := *
message := *
assume message(N,V) = w(N,V)
assume N:t <= t.iter.val(loc:iter) -> incomming_message(O,N)
\end{lstlisting}
\end{mdframed}

As a result of this architecture the program logic has to be specified twice: once in the form of invariants and another time in the actual loop. Such a program therefore has two possible origins for any error: a fault in the algorithm or an inadequate invariant. This structure also increases the difficulty of debugging, because Ivy only shows unhelpful $assert$, $assume$ and  non-determinsitic assignments ($:= *$) in its traces, rather than execution of the actual lines in the $send_messages$ function.

For the simple \textit{Step 1} this approach may still be considered viable, but for the more complex \textit{Step 2} and the failure simulation logic, the author has found that the duplication of logic and the decreased debugging efficiency makes the increased modularity achieved through iterations not worth the mental and maintenance effort that the loop invariants bring with them.

For this reason the following specification of \textit{Floodset} uses a non-modular structure where single lines predominately alter the state of the system as a whole.

\subsubsection{Cardinality relationships}
There is no cardinality operator in IVy. In set notation we would write something like $|\{n|node(n) \land failed(n)\}|$ to count the number of failed nodes in a system. When programming in a standard procedural language we would use a simple loop for this task but, as has been explained in the previous section, loops are somewhat problematic in Ivy. The author couldn't find a way to specify an inductive loop invariant for a loop counting the number of failed nodes without a cardinality operator at hand. This means that alternative encodings for the assumption $f<n$ and the statement $if rounds = f+1$ must be found.

The assumption $f<n$, ``number of failed nodes is bounded to an amount smaller than the number of total nodes'', can alternatively be expressed via $\exists n\ \neg failed(N)$.

To figure out how to alternatively encode $if rounds = f+1$ deep understanding of the algorithm is necessary. One must realise that $rounds = f+1$ entails that a ``clean round'' occurred (see \textit{Lemma 3}). Where ``clean round'' describes ``a round in which no node failed''. The processes are ready to decide after a clean round has occurred. They can decide either immediately, or an arbitrary number of rounds after the clean round. This can be modeled with non-determinism:

\begin{mdframed}[backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=15, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
clean_round_occured := (clean_round_occured | (forall N . ~node.failed_during_round(N)));
if clean_round_occured {
  decision_round := *;
};
\end{lstlisting}
\end{mdframed}

And an invariant that describes the properties of a clean round, that contains the premise of \textit{Lemma 1}:

\begin{mdframed}[backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=15, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
# lemma 1:
invariant system.clean_round_occured -> ((~node.failed(A) & ~node.failed(B)) -> (node.w(A, V) = node.w(B, V)))
\end{lstlisting}
\end{mdframed}

The (other) chosen invariants will be explained in detail later on in this paper.

\subsection{Types, functions, relations}
First the types, functions, relations and individual variables that are used throughout the program are presented:

\begin{mdframed}[backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=15, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
type node
type value
object node = {
  #graph/system related
  function val(X:node) : value
  function default(X: node) : value
  individual def : value
  relation failed(A:node)
  relation failed_before_round(A:node)
  relation failed_during_round(A:node)
  relation has_single_value(N:node)

  #state
  relation w(SELF:node, V:value)
  relation decision(SELF:node, VALUE:value)

  #communication
  relation message(N:node, V:value)
  relation incomming_message(SELF:node, OTHER:node)
  relation incomming_message_failing(SELF:node, OTHER:node)
}

object system = {
  individual decision_round:bool
  individual clean_round_occured:bool
}
\end{lstlisting}
\end{mdframed}

Their meaning will become clear as the code is presented in the following paragraphs.

A round in the specification is defined as follows:
\begin{mdframed}[backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=15, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
action round = {
  require ~node.decision(N,V) & ~decision_round;
  node.failed_before_round(N) := node.failed(N);
  call step1;
  call simulate_failure;
  clean_round_occured := (clean_round_occured | (forall N . ~node.failed_during_round(N)));
  if clean_round_occured {
    decision_round := *;
  };
  call step2;
}
export round;
\end{lstlisting}
\end{mdframed}

Line 2 makes it illegal for the environment to call \textit{round} if decisions already occurred.
Line 3 does some bookkeeping.
\textit{call step1} models the communication step. \textit{simulate\_failure} simulates failure before, during
or after \textit{step1}. After \textit{decision\_round} has been set as explained above, \textit{call step2} is
executed to model the state transition of the nodes.

\subsection{Step 1 - Message sending}
The contents of a nodes message is modeled via the relation $message(N:node, V:value)$. Where
$message(1, 0) = true; message(1, 1) = true;$ means node $1$ includes in its message the values $0$ and $1$.
Since a node never sends the same message to all nodes for any given round we don't need to include the
recipient in the relations.

The actual delivery of a message is modeled via the $incomming_message(RECIPIENT:node, SENDER:node)$ relation. Where $incoming_message(0, 1) = true; incoming_message(0, 2) = true;$ means that node $0$ receives the messages from node $1$ and node $2$.

These relations allow us to use Ivy's assignments to easily model \textit{Step 1} in a round of \textit{Floodset}.
Line 2 assigns for all nodes $A$ and values $V$ to $message(A,V)$ the boolean value of $node.w(A,V) \land \neg failed(A)$. In other words every node that hasn't failed prepares a message with all values in their $w$. Line 3 states that all nodes that haven't failed send messages to all other nodes.

\begin{mdframed}[backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=15, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
action step1 = {
  node.message(A, V) := node.w(A, V) & ~node.failed(A);
  node.incomming_message(O, A) := ~node.failed(A);
}
\end{lstlisting}
\end{mdframed}

\subsection{Step 2 - Message sending}

Line 2 of the \textit{step2} action adds to $w$ of node $N$ all values $V$ for which a node $J$ exists that has sent a message to $N$ that contains $J$.
The conditional statement in Line 4 corresponds to $if\ rounds = f+1$ in the original algorithm. Nodes with a single value in $w$ decide on that value, otherwise
they decide on the default value.
Finally, for good measure, $incomming\_message$ and $message$ are reset, but this is not necessary for the proof.

\begin{mdframed}[backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=10, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
action step2 = {
  node.w(N,V) :=
    node.w(N,V) | (exists J . node.incomming_message(N,J) &
                              node.message(J,V));

  if decision_round {

    node.has_single_value(N) :=
      (exists J . node.w(N,J)) &
      (forall V,O . O=V | (node.w(N,V) -> ~node.w(N,O)));

    node.decision(N,K) :=
      node.has_single_value(N) & node.w(N,K) & ~node.failed(N);

    node.decision(N,node.default(N)) :=
      (~node.has_single_value(N) | node.decision(N,node.default(N)))
      & ~node.failed(N);

  };

  node.incomming_message(A,B) := false;
  node.message(A,V) := false;
}
\end{lstlisting}
\end{mdframed}

\subsection{Failure simulation}

The \textit{simulate\_failure} action simulates the failure model described by Lynch.
It is called after \textit{step1} and illustrates sophisticated usage of the $assume$ functionality in Ivy.
In line 2 it choses a random number of nodes to fail. The assume in line 3 however excludes all
cases where a node that has already failed was chosen to fail in this round.
Line 4 prepares for each node $M$ a non-deterministically chosen set of nodes that it sends a message to.
Line 6 updates the \textit{incomming\_message} relation. If node $M$ failed during this round, the set of nodes
 $M$ sends messages to is changed to the values of \textit{incomming\_message\_failing}, otherwise it remains the same.
 After the nodes that \textit{failed\_during\_round} are set to \textit{failed} in line 12, line 13 assumes
 the condition $f<n$. This excludes the cases where all nodes failed.

\begin{mdframed}[backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=5, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
  action simulate_failure = {
    node.failed_during_round(N) := *;
    assume node.failed_during_round(N) -> ~node.failed(N);
    node.incomming_message_failing(N,M) := *;

    node.incomming_message(N,M) :=
      (node.failed_during_round(M) &
        node.incomming_message_failing(N,M)) |
      (~node.failed_during_round(M) &
        node.incomming_message(N,M));

    node.failed(N) := (node.failed(N) | node.failed_during_round(N));
    assume exists N . ~node.failed(N);
  }
\end{lstlisting}
\end{mdframed}

Finally, we describe the initial system state in a special \textit{after init} action. When
proofing correctness Ivy requires the initialisation to entail the invariants. All relations and functions
that are not assigned any values here are set automatically by the environment. E.g. $node.val(N)$ is
treated as a parameter the environment sets.

\begin{mdframed}[backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=5, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
after init {
  def := *;
  default(N) := def;
  w(N, V) := V=node.val(N);
  incomming_message(A, B) := false;
  decision(N, V) := false;
  clean_round_occured := false;
  decision_round := false;
  node.failed_before_round(N) := false;
  node.failed(N) := false;
  node.has_single_value(N) := false;
}
\end{lstlisting}
\end{mdframed}

\subsection{Invariants}

Now that the system has been modeled the next step of verifying correctness with Ivy is specifying inductive invariants. The simplest invariants to come up with are those that describe the correctness of the algorithm at hand. In our case these are \textit{agreement}, \textit{validity}, \textit{termination}.

\begin{mdframed}[backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=5, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
####### Correctness Conditions #######

# Agreement: No two processes decide on different values.
invariant (~node.failed(A) & ~node.failed(B)) ->
            (node.decision(A,V) = node.decision(B,V))

# every node decides on at most one value
invariant node.decision(N,V) -> (~exists O . node.decision(N,O) & O~=V)

# Validity: If all processes start with the same initial value v,
#           then v is the only possible decision value
invariant (exists N,M . node.val(N) ~= node.val(M)) |
          (system.decision_round ->
            (~node.failed(N2) -> node.decision(N2,node.val(N2))))

# Termination: All nonfaulty processes eventually decide.
invariant system.decision_round <->
            (forall N . (~node.failed(N) -> (exists V . node.decision(N,V))))
\end{lstlisting}
\end{mdframed}

If one were to run $ivy_check$ on the program without specifying any
additional invariants Ivy finds that the \textit{agreement} invariant is not inductive
and produces the a counter example to induction (CTI). One can display the CTI in a GUI.
In the left hand section of the program we can see that the system transition from state 0
to state 1 by executing \textit{round}. In the center Ivy displays state 0, meaning the state
before \textit{round} was executed. The brown edges specify the relation $w$, the purple edges
show $val$.

---SCREENSHOT 1 TODO---

By examining state 0 we realise that no node has their initial value modeled by $val$
in its $w$. The programmer must realise this is an unreachable state and exclude it by coming
up with an inductive invariant. The process of finding invariants interactively is supported by Ivy but
beyond the scope of this paper.

An invariant excluding the unreachable state is one that states that the initial value of a node is always in its $w$:
\begin{mdframed}[backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=5, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
# initial value is in W
invariant node.w(N, node.val(N))
\end{lstlisting}
\end{mdframed}

The method for finding the remaining invariants is identical:
\begin{mdframed}[backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=5, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
# the global default is the one the nodes know
invariant node.default(N) = node.def
# if a value is in a process' W, it must be initial value of some other process ( no invented values )
invariant node.w(N, V) -> (exists M . node.val(M)=V)
# lemma 1:
invariant system.clean_round_occured -> ((~node.failed(A) & ~node.failed(B)) -> (node.w(A, V) = node.w(B, V)))
\end{lstlisting}
\end{mdframed}

Since it is also required that the framework behaves correctly it makes sense to include some
correctness conditions specifically aimed at the system.

\begin{mdframed}[backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=5, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
####### system correctness checks #######

# f < n
invariant exists N . ~node.failed(N)

# 'after' a node has failed it cannot start working again
invariant node.failed_before_round(N) -> node.failed(N)

# 'after' a node has failed it will not send anything
invariant node.failed_before_round(N) -> ~node.incomming_message(M, N)

# If the decision round occured, a clean round must have occured
invariant system.decision_round -> system.clean_round_occured

\end{lstlisting}
\end{mdframed}

@@

How viable is Ivy as a tool for algorithm designers to proof correctness
of their work? The goal when turning to a tool like Ivy is certainty of
correctness. But absolute certainty is never really achieved. When specifying the
algorithm in the Ivy language one can not trivially determine that the specification
is logically equivalent to the actual algorithm. This is compounded by the fact
that the Ivy language is structurally unlike classical procedural languages. Making it challenging to achieve a one-to-one translation from a classical distributed algorithm in a language like C. An example of this is the relatively compley quantifier-logic in assignments that update the global system state.  A similar thread is posed by $assume$ statements. The author of this paper originally modeled the line

\begin{mdframed}[backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=5, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
node.w(N,V) := (node.w(N,V) | (~node.failed(N) & exists J . (node.incomming_message(N, J) & node.message(J, V))));
\end{lstlisting}
\end{mdframed}

as

\begin{mdframed}[backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=5, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
 node.tmp_w(A,V) := node.w(A, V);
    node.w(A, V) := *;
    assume (node.tmp_w(A, V) | ((node.incomming_message(A, J) & node.message(J, V)))) <-> node.w(A, V);
\end{lstlisting}
\end{mdframed}
 and did not discover that the $assume$ is bogus and excludes important possibilities from being verified. The error only became apparent when it was discovered that correctness could be proven in the absence of an invariant that should have been absolutely necessary for induction.

- an invariant can exclude too much
--todo finde so eine invariante --

Furthermore, it has already been shown that the lack of easy iteration over types makes it difficult to achieve modularity. The lack of which introduces additional uncertainty about correct modelling, especially in distributed systems.

However, as has been pointed out, it is not just the algorithm that must be correctly specified, but the underlying framework is equally important to get right. In the presented specification it is possible to remove the \textit{simulate\_failure} action and still receive a positive check from the Ivy system. This is unsurprising, as we do not have any mechanism in place that checks that the system correctly models the possibility for node failure. E.g. to check that the simulate failure action allows a node to fail and send a message only to a subset of other nodes one can purposefully insert an assertion that only fails when the desired behaviour is possible, e.g. insert
\begin{mdframed}[backgroundcolor=light-gray, roundcorner=10pt,leftmargin=1, rightmargin=1, innerleftmargin=5, innertopmargin=15,innerbottommargin=15, outerlinewidth=1, linecolor=light-gray]
\begin{lstlisting}
ensure ~(node.failed(N) &
        (exists J . ~incoming_message(J, N) &
          exists K . incomming_message(K, N) & J~=N & K~=J))
\end{lstlisting}
\end{mdframed}
after the \textit{simulate\_failure} action. If Ivy does not find a counterexample to this assert, the specification is erroneous.  In light of this, it would be beneficial to have a built in mechanism to verify the possibility of something to occur. Allowing one to conveniently verify that the modeled system allows for the possibility of a node to fail and only send messages to a subset of processes as part of the final proof.

Of course a variety of bugs and exploits do not stem from a logically unsound algorithmic foundation but implementation errors. A wonderful example is the Heartbleed exploit in OpenSSL. It would have been no help to verify the simple Heartbleed protocol in Ivy before implementing it in C, since the error was in an implementation detail allowing a buffer underflow. The point is, the implementation of the algorithm must be tested through unit testing or bounded model checkers, despite successful verification in Ivy.

- induction is quite an unintuitive for beginners to get their head around
